---
title: "Project Draft"
format: html
---

```{r}
NBA_Player_Stats <- read.csv("Datasets/2025-2026 NBA Player Stats - NBAstuffer.csv")
NBA_Player_Stats <- NBA_Player_Stats[, -1]
NBA_Player_Contracts <- read.csv("Datasets/2025-26 NBA Player Contracts.csv")
result <- merge(NBA_Player_Stats, NBA_Player_Contracts, by.x = "NAME", by.y = "Player")
head(result)
dim(result)
```

We decided to do Bayesian Inference on this question: Can we use the posterior predictive distribution to identify players who are significantly overpaid or underpaid relative to their market value?

First, we merged the two datasets we found into one dataset. Then do EDA on the dataset, explore the distribution of variables to find if there exists patterns, outliers, and we would also want to find if there's any relationship between variables which would influence our later on analysis. We would conduct the data cleaning at this step as well by dropping some columns that are unrelated to our research question and removing the outliers we find.

The two methods we decided to use are:

1.  Using stan to build a Bayesian linear regression model.

    Before building the model, we will do variable selection based off the definition of the variables, and using methods like VIFs, forward model selection to choose the appropriate variables for the model. Then we will use the model to do stan, choosing the appropriate prior and posterior for the variables, and do posterior predictive distribution to identify players who are significantly overpaid or underpaid relative to their market value.

2.  MCMC Metropolis-Hastings

    We decided to do something related to MH algorithm. We implemented the Metropolis-Hastings algorithm manually to sample from the posterior distribution. Then, compare the MCMC result with the Stan method result. This would determine whether or not those two methods produce consistent posterior estimates.

## EDA

```{r}
# packages being used
library(tidyverse)
# library(GGally)
library(corrplot)
library(dplyr)


```

We notice that, our contracted salary(X2025.26 to X2030.31) are treated to be character instead of numeric because of the extra dollar sign(\$). We must remove the sign and set these to numeric for later analysis.

```{r}
result <- result %>%
  mutate(across(starts_with("X20"), ~ as.numeric(gsub("\\$", "", .x))))
class(result$X2025.26)
```

We also need to check whether there is any NA values for our response variable, if NA exists we would need to drop that row.

```{r}
result[is.na(result$X2025.26), ]
```

There exists one row which misses the value for our response variable, we decide to drop.

```{r}
result_final <- result[!is.na(result$X2025.26), ]
dim(result_final)
```




two models - one with prior less information, one prior information

linear regression

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(GGally)

NBA_Player_Stats_clean <- result[, c("NAME", "TEAM", "CUR", "POS", "AGE", "GP", "MpG", "USG.", "TO.", "TS.", "PpG", "RpG", "ApG", "SpG", "BpG","X2025.26")]
# head(NBA_Player_Stats_clean)

df <- NBA_Player_Stats_clean 
df <- rename(df, Salary = X2025.26)

# Original Salary
p1 <- ggplot(df, aes(x = Salary)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_density(color = "steelblue") +
  labs(title = "Contracted Salary Distribution(2025-2026)")

p1

# Log Salary
p2 <- ggplot(df, aes(x = log(Salary))) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_density(color = "coral") +
  labs(title = "Log Salary Distribution(2025-2026)")

p2

# Position
ggplot(df, aes(x = POS, y = log(Salary))) +
  geom_boxplot() + 
  labs(title = "Side by side boxplot of Log(Salary) by position")
```
We notice that the distribution of the contracted salary(X2025.26) is heavily right-skewed. A log transformation is necessary in order to standardize the variable.

```{r, warning=FALSE, message=FALSE}
pred_vars <- c("USG.", "TO.", "TS.", "PpG", "RpG", "ApG", "SpG", "BpG")

# Histogram
df %>%
  select(all_of(pred_vars)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 25) +
  facet_wrap(~name, scales = "free") +
  labs(title = "Predictor distributions")

# Variables vs log(Salary) Scatterplot
df %>%
  mutate(log_salary = log(Salary)) %>%
  select(log_salary, all_of(pred_vars)) %>%
  pivot_longer(-log_salary) %>%
  ggplot(aes(x = value, y = log_salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "Predictors vs log(Salary)")
```

```{r, warning=FALSE, message=FALSE}
# Correlation Plot / Matrix
cor_mat <- cor(df[, pred_vars], use = "complete.obs")
corrplot::corrplot(cor_mat, method = "color", addCoef.col = "black", 
                   number.cex = 0.7, tl.cex = 0.8)

# Pairs plot（log Salary）
df %>%
  mutate(log_salary = log(Salary)) %>%
  select(log_salary, all_of(pred_vars)) %>%
  ggpairs()

# VIF
ols_fit <- lm(log(Salary) ~ USG. + TO. + TS. + PpG + RpG + ApG + SpG + BpG,
              data = df)
vif(ols_fit)
```
We notice that, for the heat map, there's no correlation coefficient greater than $0.8$, which means that all the variables at this stage are not considered to be highly correlated to each other. So we decide not to drop any of them.

Note that for our VIF values, there exists one VIF value of $5.983835$(`PpG`) which is normally considered to be a concerning threshold that requires further investigation since this variable could potentially affecting coefficient estimates. Since `PpG` represents points per game, which means higher `PpG` surely results in higher `Salary`. But here we decide not to drop this column, since this VIF value is not extremely high(less than 10), and this variable acts like a key factor to consider for our research question.


USG.      TO.      TS.      PpG      RpG      ApG      SpG      BpG 
## Model 1:
Our naive model is linear regression.
$$\log(salary)_i \sim \mathcal{N}(\alpha + \beta_1 \cdot USG_i + \beta_2\cdot TO_i\cdot + \beta_3\cdot TS_i + \beta_4 \cdot PpG + \beta_5 \cdot RpG_i + \beta_6 \cdot ApG_i + \beta_7 \cdot SpG_i +\beta_8 \cdot BpG_i, σ^2)$$

And we have $$\alpha \sim \mathcal{N} (16, 2) \\ \beta_1 \sim \mathcal{N} (), \\ \beta_2 \sim \mathcal{N} (), \\  \beta_3 \sim \mathcal{N} (), \\ \beta_4 \sim \mathcal{N} (), \\ \beta_5 \sim \mathcal{N} (), \\ \beta_6 \sim \mathcal{N} (), \\ \beta_7 \sim \mathcal{N} (), \\ \beta_8 \sim \mathcal{N} (), \\ \sigma \sim ()$$. 




## Model 2:
Our complex model is Hierarchical Linear Regression (layered by positions)
$$\log(salary)_i \sim \mathcal{N}(\alpha_j + \beta_1 \cdot USG_i + \beta_2\cdot TO_i\cdot + \beta_3\cdot TS_i + \beta_4 \cdot PpG + 
\beta_5 \cdot RpG_i + \beta_6 \cdot ApG_i + \beta_7 \cdot SpG_i +\beta_8 \cdot BpG_i, σ^2)$$

And we have $$\alpha_j \sim \mathcal{N} (\mu_a, \tau^2) \\ \beta_1 \sim \mathcal{N} (), \\ \beta_2 \sim \mathcal{N} (), \\  \beta_3 \sim \mathcal{N} (), \\ \beta_4 \sim \mathcal{N} (), \\ \beta_5 \sim \mathcal{N} (), \\ \beta_6 \sim \mathcal{N} (), \\ \beta_7 \sim \mathcal{N} (), \\ \beta_8 \sim \mathcal{N} () \\ \tau \sim ()$$.


