---
title: "Project Draft"
format: html
---

## Data Overview

```{r, warning=FALSE, message=FALSE}
# packages being used
library(tidyverse)
library(corrplot)
library(dplyr)
library(GGally)
library(car)

NBA_Player_Stats <- read.csv("Datasets/2025-2026 NBA Player Stats - NBAstuffer.csv")
NBA_Player_Stats <- NBA_Player_Stats[, -1]
NBA_Player_Contracts <- read.csv("Datasets/2025-26 NBA Player Contracts.csv")
result <- merge(NBA_Player_Stats, NBA_Player_Contracts, by.x = "NAME", by.y = "Player")
head(result)
dim(result)
```

We decided to do Bayesian Inference on this question: Can we use the posterior predictive distribution to identify players who are significantly overpaid or underpaid relative to their market value?

First, we merged the two datasets we found into one dataset. Then we perform EDA on the dataset, explore the distribution of variables to find if there exists patterns, outliers, and we would also want to find if there's any relationship between variables which would influence our later on analysis. Data cleaning is also conducted at this step as well by dropping some columns that are unrelated to our research question and removing the outliers we find.

The two methods we decided to use are:

1.  Using stan to build a Bayesian linear regression model.

    Before building the model, we will do variable selection based off the definition of the variables, and using methods like VIFs, forward model selection to choose the appropriate variables for the model. Then we will use the model to do stan, choosing the appropriate prior and posterior for the variables, and do posterior predictive distribution to identify players who are significantly overpaid or underpaid relative to their market value.

2.  MCMC Metropolis-Hastings

    We decided to do something related to MH algorithm. We implemented the Metropolis-Hastings algorithm manually to sample from the posterior distribution. Then, compare the MCMC result with the Stan method result. This allows us to determine whether or not those two methods produce consistent posterior estimates.

## Data Cleaning

We notice that our contracted salary columns (X2025.26 to X2030.31) are treated as character instead of numeric because of the extra dollar sign (\$). We must remove the sign and set these to numeric for later analysis.

```{r}
result <- result %>%
  mutate(across(starts_with("X20"), ~ as.numeric(gsub("\\$", "", .x))))
class(result$X2025.26)
```

We also need to check whether there are any NA values for our response variable; if NA exists we would need to drop that row.

```{r}
result[is.na(result$X2025.26), ]
```

There exists one row which misses the value for our response variable, we drop this observation.

```{r}
result_final <- result[!is.na(result$X2025.26), ]
dim(result_final)
```

Next, we select the relevant columns for our research question and rename them for clarity.

```{r}
NBA_Player_Stats_clean <- result_final[, c("NAME", "TEAM", "CUR", "POS", "AGE", "GP", "MpG", "USG.", "TO.", "TS.", "PpG", "RpG", "ApG", "SpG", "BpG", "X2025.26")]

df <- NBA_Player_Stats_clean
df <- rename(df, Salary = X2025.26)   # rename salary
names(df) <- sub("\\..*", "", names(df))  # remove dots in names
head(df)
```

## EDA

### Salary Distribution

```{r, warning=FALSE, message=FALSE}
# Original Salary
p1 <- ggplot(df, aes(x = Salary)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_density(color = "steelblue") +
  labs(title = "Contracted Salary Distribution (2025-2026)")

p1

# Log Salary
p2 <- ggplot(df, aes(x = log(Salary))) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_density(color = "coral") +
  labs(title = "Log Salary Distribution (2025-2026)")

p2
```

We notice that the distribution of the contracted salary (X2025.26) is heavily right-skewed. A log transformation is necessary in order to standardize the variable.

### Salary by Position

```{r, warning=FALSE, message=FALSE}
# Position
ggplot(df, aes(x = POS, y = log(Salary))) +
  geom_boxplot() +
  labs(title = "Side by side boxplot of Log(Salary) by position")
```

### Outliers Removal

```{r}
# remove outliers
df <- df %>%
  group_by(POS) %>%
  mutate(
    Q1 = quantile(log(Salary), 0.25, na.rm = TRUE),
    Q3 = quantile(log(Salary), 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    lower = Q1 - 1.5 * IQR
  ) %>%
  filter(!(POS %in% c("F", "G") & log(Salary) < lower)) %>%
  select(-Q1, -Q3, -IQR, -lower) %>%
  ungroup()
dim(df)
```

```{r, warning=FALSE, message=FALSE}
# Position boxplot after data cleaning
ggplot(df, aes(x = POS, y = log(Salary))) +
  geom_boxplot() +
  labs(title = "Side by side boxplot of Log(Salary) by position (after cleaning)")
```

### Predictor Distributions and Relationships with Log(Salary)

```{r, warning=FALSE, message=FALSE}
pred_vars <- c("USG", "TO", "TS", "PpG", "RpG", "ApG", "SpG", "BpG")

# Histogram
df %>%
  select(all_of(pred_vars)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 25) +
  facet_wrap(~name, scales = "free") +
  labs(title = "Predictor distributions")

# Variables vs log(Salary) Scatterplot
df %>%
  mutate(log_salary = log(Salary)) %>%
  select(log_salary, all_of(pred_vars)) %>%
  pivot_longer(-log_salary) %>%
  ggplot(aes(x = value, y = log_salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "Predictors vs log(Salary)")
```

### Correlation and Multicollinearity Check

```{r, warning=FALSE, message=FALSE}
# Correlation Plot / Matrix
cor_mat <- cor(df[, pred_vars], use = "complete.obs")
corrplot::corrplot(cor_mat, method = "color", addCoef.col = "black",
                   number.cex = 0.7, tl.cex = 0.8)

# Pairs plot (log Salary)
df %>%
  mutate(log_salary = log(Salary)) %>%
  select(log_salary, all_of(pred_vars)) %>%
  ggpairs()

# VIF
ols_fit <- lm(log(Salary) ~ USG + TO + TS + PpG + RpG + ApG + SpG + BpG,
              data = df)
vif(ols_fit)
```

We notice that for the heat map, there's no correlation coefficient greater than $0.8$, which means that all the variables at this stage are not considered to be highly correlated to each other. So we decide not to drop any of them.

Note that for our VIF values, there exists one VIF value of $6.016229$ (`PpG`) which is normally considered to be a concerning threshold that requires further investigation since this variable could potentially be affecting coefficient estimates. Since `PpG` represents points per game, higher `PpG` surely results in higher `Salary`. However, we decide not to drop this column, since this VIF value is not extremely high (less than 10), and this variable acts as a key factor to consider for our research question.

According to the distributions of our explanatory variables, we notice that those variables all have different ranges. In order to do later on analysis, we need to standardize our 8 predictors to make our prior specifications become consistent. Without standardization, coefficients are not comparable. In conclusion, Standardization ensures that the same prior is equally informative for all predictors regardless of their original units, and the posterior coefficients can be directly compared to assess the relative importance of each stat.

```{r}
df <- df %>%
  mutate(
    log_salary = log(Salary),
    USG_s = scale(USG)[,1],
    TO_s  = scale(TO)[,1],
    TS_s  = scale(TS)[,1],
    PpG_s = scale(PpG)[,1],
    RpG_s = scale(RpG)[,1],
    ApG_s = scale(ApG)[,1],
    SpG_s = scale(SpG)[,1],
    BpG_s = scale(BpG)[,1]
  )
head(df)
```

## Model 1: Bayesian Linear Regression

Our naive model is linear regression.

$$\log(salary)_i \sim \mathcal{N}(\alpha + \beta_1 \cdot USG_i + \beta_2\cdot TO_i + \beta_3\cdot TS_i + \beta_4 \cdot PpG_i + \beta_5 \cdot RpG_i + \beta_6 \cdot ApG_i + \beta_7 \cdot SpG_i +\beta_8 \cdot BpG_i,\ \sigma^2)$$

And we have

$$\alpha \sim \mathcal{N}(16, 2) \\ \beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8 \sim \mathcal{N}(0,1) \\ \sigma \sim \text{Half-Normal}(0, 1)$$

```{r}
## for checking
ols_fit <- lm(log(Salary) ~ USG_s + TO_s + TS_s + PpG_s + 
                RpG_s + ApG_s + SpG_s + BpG_s, data = df)
summary(ols_fit)$sigma
```

## Model 2: Hierarchical Bayesian Linear Regression

Our complex model is Hierarchical Linear Regression with position-specific intercepts.

$$\log(salary)_i \sim \mathcal{N}(\alpha_{j_[i]} + \beta_1 \cdot USG_i + \beta_2\cdot TO_i + \beta_3\cdot TS_i + \beta_4 \cdot PpG_i + \beta_5 \cdot RpG_i + \beta_6 \cdot ApG_i + \beta_7 \cdot SpG_i +\beta_8 \cdot BpG_i,\ \sigma^2)$$ Where $j[i]$ represents the i-th basketball player's position, $j = 1,2,...,7$, that we have different 7 positions in total.

$$\alpha_j \sim \mathcal{N}(\mu_\alpha, \tau^2), \ j = 1,2,...,7 \\ \mu_\alpha \sim \mathcal{N}(16,2) \\ \tau \sim \text{Half-Normal}(0, 1) \\ \beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8 \sim \mathcal{N}(0,1) \\ \sigma \sim \text{Half-Normal}(0, 1)$$

## Run Posterior Using Stan

```{r, warning=FALSE, message=FALSE}
library(rstan)
library(loo)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

pred_vars <- c("USG", "TO", "TS", "PpG", "RpG", "ApG", "SpG", "BpG")

# standardize 
X_raw <- df[, pred_vars]
X_scaled <- scale(X_raw)

y <- log(df$Salary)

# position index
pos_levels <- levels(factor(df$POS))
pos_id     <- as.integer(factor(df$POS, levels = pos_levels))
J          <- length(pos_levels)
N          <- nrow(df)
K          <- length(pred_vars)

cat("N =", N, "| K =", K, "| J =", J, "\n")
cat("Positions:", pos_levels, "\n")

# model 1 data
stan_data_m1 <- list(
  N = N,
  K = K,
  X = X_scaled,
  y = y
)

fit_m1 <- stan(
  file    = "model1.stan",
  data    = stan_data_m1,
  iter    = 4000,          # 2000 warmup + 2000 sampling
  warmup  = 2000,
  chains  = 4,
  seed    = 42,
  control = list(adapt_delta = 0.95),
  refresh = 0
)

print(fit_m1, pars = c("alpha", "beta", "sigma"), probs = c(0.025, 0.5, 0.975))

# model 2 data
stan_data_m2 <- list(
  N      = N,
  K      = K,
  J      = J,
  X      = X_scaled,
  y      = y,
  pos_id = pos_id
)

fit_m2 <- stan(
  file    = "model2.stan",
  data    = stan_data_m2,
  iter    = 4000,
  warmup  = 2000,
  chains  = 4,
  seed    = 42,
  control = list(adapt_delta = 0.95),
  refresh = 0
)

print(fit_m2, pars = c("mu_a", "tau", "alpha", "beta", "sigma"),
      probs = c(0.025, 0.5, 0.975))

stan_diag_m1 <- summary(fit_m1)$summary
stan_diag_m2 <- summary(fit_m2)$summary

cat("\nModel 1 - max R-hat:", max(stan_diag_m1[, "Rhat"], na.rm = TRUE), "\n")
cat("Model 2 - max R-hat:", max(stan_diag_m2[, "Rhat"], na.rm = TRUE), "\n")
```

## LOO-CV for Result Comparison

```{r, warning=FALSE, message=FALSE}
# leave one out cross validation
log_lik_m1 <- extract_log_lik(fit_m1, merge_chains = FALSE)
log_lik_m2 <- extract_log_lik(fit_m2, merge_chains = FALSE)

loo_m1 <- loo(log_lik_m1, r_eff = relative_eff(exp(log_lik_m1)))
loo_m2 <- loo(log_lik_m2, r_eff = relative_eff(exp(log_lik_m2)))

cat("\nLoo compare:", loo_compare(loo_m1, loo_m2), "\n")
```

## 95% Posterior Predictive Interval of Results

```{r, warning=FALSE, message=FALSE}
# Model 1
y_rep_model1  <- extract(fit_m1, "y_rep")$y_rep

# posterior predictive distribution mean & 95% ci（log scale）
ppd_model1_mean  <- apply(y_rep_model1, 2, mean)
ppd_model1_lower <- apply(y_rep_model1, 2, quantile, 0.025)
ppd_model1_upper <- apply(y_rep_model1, 2, quantile, 0.975)

results_model1 <- data.frame(
  NAME          = df$NAME,
  POS           = df$POS,
  Actual_Salary = df$Salary,
  log_Salary    = y,
  PPD_mean      = ppd_model1_mean,
  PPD_lower     = ppd_model1_lower,
  PPD_upper     = ppd_model1_upper,
  # change back to original salary
  Predicted_Salary = exp(ppd_model1_mean),
  Lower_Salary     = exp(ppd_model1_lower),
  Upper_Salary     = exp(ppd_model1_upper)
)

# overpaid / underpaid
results_model1$Status <- "Fair"
results_model1$Status[results_model1$log_Salary > results_model1$PPD_upper] <- "Overpaid"
results_model1$Status[results_model1$log_Salary < results_model1$PPD_lower] <- "Underpaid"

cat("\nOverpaid players:\n")
print(results_model1[results_model1$Status == "Overpaid",
              c("NAME", "POS", "Actual_Salary", "Predicted_Salary")])

cat("\nUnderpaid players:\n")
print(results_model1[results_model1$Status == "Underpaid",
              c("NAME", "POS", "Actual_Salary", "Predicted_Salary")])


# Model 2
y_rep_model2  <- extract(fit_m2, "y_rep")$y_rep

# posterior predictive distribution mean & 95% ci（log scale）
ppd_model2_mean  <- apply(y_rep_model2, 2, mean)
ppd_model2_lower <- apply(y_rep_model2, 2, quantile, 0.025)
ppd_model2_upper <- apply(y_rep_model2, 2, quantile, 0.975)

results_model2 <- data.frame(
  NAME          = df$NAME,
  POS           = df$POS,
  Actual_Salary = df$Salary,
  log_Salary    = y,
  PPD_mean      = ppd_model2_mean,
  PPD_lower     = ppd_model2_lower,
  PPD_upper     = ppd_model2_upper,
  # change back to original salary
  Predicted_Salary = exp(ppd_model2_mean),
  Lower_Salary     = exp(ppd_model2_lower),
  Upper_Salary     = exp(ppd_model2_upper)
)

# overpaid / underpaid
results_model2$Status <- "Fair"
results_model2$Status[results_model2$log_Salary > results_model2$PPD_upper] <- "Overpaid"
results_model2$Status[results_model2$log_Salary < results_model2$PPD_lower] <- "Underpaid"

cat("\nOverpaid players:\n")
print(results_model2[results_model2$Status == "Overpaid",
              c("NAME", "POS", "Actual_Salary", "Predicted_Salary")])

cat("\nUnderpaid players:\n")
print(results_model2[results_model2$Status == "Underpaid",
              c("NAME", "POS", "Actual_Salary", "Predicted_Salary")])
```

### True salaries against the 95% posterior predictive intervals Plot

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(scales)

make_plot_df <- function(res, model_label) {
  res %>%
    arrange(Actual_Salary) %>%
    mutate(
      player_idx = row_number(),
      Model      = model_label,
      Status     = factor(Status, levels = c("Underpaid", "Fair", "Overpaid"))
    )
}

plot_df1 <- make_plot_df(results_model1, "Model 1")
plot_df2 <- make_plot_df(results_model2, "Model 2")
plot_df  <- bind_rows(plot_df1, plot_df2)

status_colours <- c(
  Underpaid = "#2166ac",
  Fair      = "#b0b0b0",
  Overpaid  = "#d73027"
)

salary_cap <- quantile(plot_df$Upper_Salary, 0.97) / 1e6  # clip top 3%

ggplot(plot_df, aes(x = player_idx)) +

  # 95 % posterior predictive interval ribbon
  geom_ribbon(
    aes(
      ymin = pmax(Lower_Salary / 1e6, 0.01),
      ymax = pmin(Upper_Salary / 1e6, salary_cap)
    ),
    fill = "#a8c8e8", alpha = 0.45
  ) +

  # Predicted salary line
  geom_line(
    aes(y = pmin(Predicted_Salary / 1e6, salary_cap)),
    colour = "#4393c3", linewidth = 0.6, linetype = "dashed"
  ) +

  # Actual salary dots, coloured by pay status
  geom_point(
    aes(y = Actual_Salary / 1e6, colour = Status, size = Status),
    alpha = 0.9
  ) +

  scale_colour_manual(values = status_colours, name = "Pay Status") +
  scale_size_manual(
    values = c(Underpaid = 2.2, Fair = 1.2, Overpaid = 2.8),
    guide = "none"
  ) +

  scale_y_log10(
    labels = function(x) paste0("$", round(x, 1), "M"),
    breaks = c(0.5, 1, 2, 5, 10, 20, 40),
    limits = c(0.3, max(plot_df$Actual_Salary / 1e6) * 1.1)
  ) +

  facet_wrap(~Model, ncol = 1) +

  labs(
    title    = "True Salaries vs 95% Posterior Predictive Intervals",
    subtitle = "Log scale · sorted by actual salary · blue band = 95% CI · dashed = posterior mean",
    x        = "Player rank (by actual salary)",
    y        = "Salary (USD millions, log scale)"
  ) +

  theme_bw(base_size = 13) +
  theme(
    legend.position   = "top",
    legend.key.size   = unit(1.2, "lines"),
    strip.background  = element_rect(fill = "#e8eef4"),
    strip.text        = element_text(face = "bold", size = 12),
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(colour = "grey90"),
    plot.title        = element_text(face = "bold", size = 14),
    plot.subtitle     = element_text(colour = "grey40", size = 10)
  )
```

## Variational Inference

```{r, warning=FALSE, message=FALSE}
# using model1.stan for VI
fit_m1_vi <- vb(
  stan_model("model1.stan"),
  data      = stan_data_m1,
  algorithm = "meanfield",
  seed      = 42,
  refresh = 0
)

# using model2.stan for VI
fit_m2_vi <- vb(
  stan_model("model2.stan"),
  data      = stan_data_m2,
  algorithm = "meanfield",
  seed      = 42,
  refresh = 0
)
```

### Model 1 HMC and VI Comparison plot

```{r, warning=FALSE, message=FALSE}
library(bayesplot)
library(tidyverse)

pars_of_interest <- c(paste0("beta[", 1:8, "]"), "sigma")

draws_hmc <- as.matrix(fit_m1,    pars = pars_of_interest)
draws_vi  <- as.matrix(fit_m1_vi, pars = pars_of_interest)

beta_labels <- c("USG","TO","TS","PpG","RpG","ApG","SpG","BpG","sigma")

# HMC vs VI的posterior density
df_hmc <- as.data.frame(draws_hmc) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "HMC")

df_vi <- as.data.frame(draws_vi) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "VI (Mean-field)")

df_combined <- bind_rows(df_hmc, df_vi)

param_map <- setNames(beta_labels, pars_of_interest)
df_combined$parameter <- param_map[df_combined$parameter]
df_combined$parameter <- factor(df_combined$parameter, levels = beta_labels)

ggplot(df_combined, aes(x = value, fill = method, color = method)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free", ncol = 3) +
  scale_fill_manual(values  = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  scale_color_manual(values = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  labs(
    title    = "Posterior distributions: HMC vs VI (Model 1)",
    subtitle = "VI tends to underestimate posterior variance (narrower curves)",
    x = "Parameter value", y = "Density", fill = "Method", color = "Method"
  ) +
  theme_bw() +
  theme(legend.position = "top")

# quantify the mean of posterior and difference between SD
comparison_df <- data.frame(
  Parameter = beta_labels,
  HMC_mean  = colMeans(draws_hmc),
  VI_mean   = colMeans(draws_vi),
  HMC_sd    = apply(draws_hmc, 2, sd),
  VI_sd     = apply(draws_vi,  2, sd)
) %>%
  mutate(
    mean_diff  = round(abs(HMC_mean - VI_mean), 4),
    sd_ratio   = round(VI_sd / HMC_sd, 3),       # <1 represents VI undermines variance
    VI_narrower = ifelse(sd_ratio < 1, "Yes", "No")
  )

print(comparison_df)

# visualize SD comparison
comparison_df %>%
  filter(Parameter != "sigma") %>%  
  select(Parameter, HMC_sd, VI_sd) %>%
  pivot_longer(-Parameter, names_to = "Method", values_to = "SD") %>%
  mutate(Method = ifelse(Method == "HMC_sd", "HMC", "VI")) %>%
  ggplot(aes(x = Parameter, y = SD, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("HMC" = "#3266ad", "VI" = "#D85A30")) +
  labs(
    title    = "Posterior standard deviation: HMC vs VI",
    subtitle = "VI consistently underestimates uncertainty (lower SD)",
    x = "Parameter", y = "Posterior SD"
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

### Model 2 HMC and VI Comparison plot

```{r, warning=FALSE, message=FALSE}

# extract posterior samples
pars_beta  <- paste0("beta[",  1:8, "]")
pars_alpha <- paste0("alpha[", 1:7, "]")
pars_hyper <- c("mu_a", "tau", "sigma")

pos_labels  <- pos_levels # 7 positions in total
beta_labels <- c("USG","TO","TS","PpG","RpG","ApG","SpG","BpG")

# comparison for Beta coefficient（HMC vs VI)
draws_hmc_beta <- as.matrix(fit_m2,    pars = pars_beta)
draws_vi_beta  <- as.matrix(fit_m2_vi, pars = pars_beta)

df_hmc_beta <- as.data.frame(draws_hmc_beta) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "HMC")

df_vi_beta <- as.data.frame(draws_vi_beta) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "VI (Mean-field)")

param_map_beta <- setNames(beta_labels, pars_beta)

bind_rows(df_hmc_beta, df_vi_beta) %>%
  mutate(parameter = param_map_beta[parameter],
         parameter = factor(parameter, levels = beta_labels)) %>%
  ggplot(aes(x = value, fill = method, color = method)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free", ncol = 4) +
  scale_fill_manual(values  = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  scale_color_manual(values = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  labs(
    title    = "Model 2: Beta posteriors — HMC vs VI",
    subtitle = "Shared coefficients across all positions",
    x = "Parameter value", y = "Density",
    fill = "Method", color = "Method"
  ) +
  theme_bw() +
  theme(legend.position = "top")

# compare: Position intercepts alpha[j]  ─────────────
draws_hmc_alpha <- as.matrix(fit_m2,    pars = pars_alpha)
draws_vi_alpha  <- as.matrix(fit_m2_vi, pars = pars_alpha)

param_map_alpha <- setNames(pos_labels, pars_alpha)

df_hmc_alpha <- as.data.frame(draws_hmc_alpha) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "HMC")

df_vi_alpha <- as.data.frame(draws_vi_alpha) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "VI (Mean-field)")

bind_rows(df_hmc_alpha, df_vi_alpha) %>%
  mutate(parameter = param_map_alpha[parameter],
         parameter = factor(parameter, levels = pos_labels)) %>%
  ggplot(aes(x = value, fill = method, color = method)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free", ncol = 4) +
  scale_fill_manual(values  = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  scale_color_manual(values = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  labs(
    title    = "Model 2: Position intercepts α_j — HMC vs VI",
    subtitle = "Each position has its own intercept — key feature of hierarchical model",
    x = "Parameter value", y = "Density",
    fill = "Method", color = "Method"
  ) +
  theme_bw() +
  theme(legend.position = "top")

# compare Hyperparameters: mu_a, tau, sigma 
draws_hmc_hyper <- as.matrix(fit_m2,    pars = pars_hyper)
draws_vi_hyper  <- as.matrix(fit_m2_vi, pars = pars_hyper)

df_hmc_hyper <- as.data.frame(draws_hmc_hyper) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "HMC")

df_vi_hyper <- as.data.frame(draws_vi_hyper) %>%
  pivot_longer(everything(), names_to = "parameter", values_to = "value") %>%
  mutate(method = "VI (Mean-field)")

bind_rows(df_hmc_hyper, df_vi_hyper) %>%
  mutate(parameter = factor(parameter, levels = pars_hyper)) %>%
  ggplot(aes(x = value, fill = method, color = method)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free", ncol = 3) +
  scale_fill_manual(values  = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  scale_color_manual(values = c("HMC" = "#3266ad", "VI (Mean-field)" = "#D85A30")) +
  labs(
    title    = "Model 2: Hyperparameters — HMC vs VI",
    subtitle = "mu_a = global mean intercept, tau = between-position SD",
    x = "Parameter value", y = "Density",
    fill = "Method", color = "Method"
  ) +
  theme_bw() +
  theme(legend.position = "top")

# Quantify the SD comparison for all parameters
all_pars   <- c(pars_beta, pars_alpha, pars_hyper)
all_labels <- c(beta_labels, pos_labels, pars_hyper)

draws_hmc_all <- as.matrix(fit_m2,    pars = all_pars)
draws_vi_all  <- as.matrix(fit_m2_vi, pars = all_pars)

comparison_m2 <- data.frame(
  Parameter = all_labels,
  Type      = c(rep("Beta", 8), rep("Alpha (position)", 7), rep("Hyperparameter", 3)),
  HMC_mean  = colMeans(draws_hmc_all),
  VI_mean   = colMeans(draws_vi_all),
  HMC_sd    = apply(draws_hmc_all, 2, sd),
  VI_sd     = apply(draws_vi_all,  2, sd)
) %>%
  mutate(
    sd_ratio    = round(VI_sd / HMC_sd, 3),
    VI_narrower = ifelse(sd_ratio < 1, "Yes", "No")
  )

print(comparison_m2)

# SD comparison visualization by bar chart(differs in position)
comparison_m2 %>%
  select(Parameter, Type, HMC_sd, VI_sd) %>%
  pivot_longer(c(HMC_sd, VI_sd), names_to = "Method", values_to = "SD") %>%
  mutate(Method = ifelse(Method == "HMC_sd", "HMC", "VI")) %>%
  ggplot(aes(x = Parameter, y = SD, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  facet_wrap(~Type, scales = "free_x", ncol = 1) +
  scale_fill_manual(values = c("HMC" = "#3266ad", "VI" = "#D85A30")) +
  labs(
    title    = "Model 2: Posterior SD comparison — HMC vs VI",
    subtitle = "VI consistently produces narrower posteriors across all parameter types",
    x = NULL, y = "Posterior SD"
  ) +
  theme_bw() +
  theme(
    legend.position = "top",
    axis.text.x     = element_text(angle = 45, hjust = 1)
  )
```
