---
title: "Project Draft"
format: html
---

## Data Overview

```{r, warning=FALSE, message=FALSE}
# packages being used
library(tidyverse)
library(corrplot)
library(dplyr)
library(GGally)
library(car)

NBA_Player_Stats <- read.csv("Datasets/2025-2026 NBA Player Stats - NBAstuffer.csv")
NBA_Player_Stats <- NBA_Player_Stats[, -1]
NBA_Player_Contracts <- read.csv("Datasets/2025-26 NBA Player Contracts.csv")
result <- merge(NBA_Player_Stats, NBA_Player_Contracts, by.x = "NAME", by.y = "Player")
head(result)
dim(result)
```

We decided to do Bayesian Inference on this question: Can we use the posterior predictive distribution to identify players who are significantly overpaid or underpaid relative to their market value?

First, we merged the two datasets we found into one dataset. Then we perform EDA on the dataset, explore the distribution of variables to find if there exists patterns, outliers, and we would also want to find if there's any relationship between variables which would influence our later on analysis. Data cleaning is also conducted at this step as well by dropping some columns that are unrelated to our research question and removing the outliers we find.

The two methods we decided to use are:

1.  Using stan to build a Bayesian linear regression model.

    Before building the model, we will do variable selection based off the definition of the variables, and using methods like VIFs, forward model selection to choose the appropriate variables for the model. Then we will use the model to do stan, choosing the appropriate prior and posterior for the variables, and do posterior predictive distribution to identify players who are significantly overpaid or underpaid relative to their market value.

2.  MCMC Metropolis-Hastings

    We decided to do something related to MH algorithm. We implemented the Metropolis-Hastings algorithm manually to sample from the posterior distribution. Then, compare the MCMC result with the Stan method result. This allows us to determine whether or not those two methods produce consistent posterior estimates.

## Data Cleaning

We notice that our contracted salary columns (X2025.26 to X2030.31) are treated as character instead of numeric because of the extra dollar sign (\$). We must remove the sign and set these to numeric for later analysis.

```{r}
result <- result %>%
  mutate(across(starts_with("X20"), ~ as.numeric(gsub("\\$", "", .x))))
class(result$X2025.26)
```

We also need to check whether there are any NA values for our response variable; if NA exists we would need to drop that row.

```{r}
result[is.na(result$X2025.26), ]
```

There exists one row which misses the value for our response variable, we drop this observation.

```{r}
result_final <- result[!is.na(result$X2025.26), ]
dim(result_final)
```

Next, we select the relevant columns for our research question and rename them for clarity.

```{r}
NBA_Player_Stats_clean <- result_final[, c("NAME", "TEAM", "CUR", "POS", "AGE", "GP", "MpG", "USG.", "TO.", "TS.", "PpG", "RpG", "ApG", "SpG", "BpG", "X2025.26")]

df <- NBA_Player_Stats_clean
df <- rename(df, Salary = X2025.26)   # rename salary
names(df) <- sub("\\..*", "", names(df))  # remove dots in names
head(df)
```

## EDA

### Salary Distribution

```{r, warning=FALSE, message=FALSE}
# Original Salary
p1 <- ggplot(df, aes(x = Salary)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_density(color = "steelblue") +
  labs(title = "Contracted Salary Distribution (2025-2026)")

p1

# Log Salary
p2 <- ggplot(df, aes(x = log(Salary))) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_density(color = "coral") +
  labs(title = "Log Salary Distribution (2025-2026)")

p2
```

We notice that the distribution of the contracted salary (X2025.26) is heavily right-skewed. A log transformation is necessary in order to standardize the variable.

### Salary by Position

```{r, warning=FALSE, message=FALSE}
# Position
ggplot(df, aes(x = POS, y = log(Salary))) +
  geom_boxplot() +
  labs(title = "Side by side boxplot of Log(Salary) by position")
```

### Outliers Removal

```{r}
# remove outliers
df <- df %>%
  group_by(POS) %>%
  mutate(
    Q1 = quantile(log(Salary), 0.25, na.rm = TRUE),
    Q3 = quantile(log(Salary), 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    lower = Q1 - 1.5 * IQR
  ) %>%
  filter(!(POS %in% c("F", "G") & log(Salary) < lower)) %>%
  select(-Q1, -Q3, -IQR, -lower) %>%
  ungroup()
dim(df)
```

```{r, warning=FALSE, message=FALSE}
# Position boxplot after data cleaning
ggplot(df, aes(x = POS, y = log(Salary))) +
  geom_boxplot() +
  labs(title = "Side by side boxplot of Log(Salary) by position (after cleaning)")
```

### Predictor Distributions and Relationships with Log(Salary)

```{r, warning=FALSE, message=FALSE}
pred_vars <- c("USG", "TO", "TS", "PpG", "RpG", "ApG", "SpG", "BpG")

# Histogram
df %>%
  select(all_of(pred_vars)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 25) +
  facet_wrap(~name, scales = "free") +
  labs(title = "Predictor distributions")

# Variables vs log(Salary) Scatterplot
df %>%
  mutate(log_salary = log(Salary)) %>%
  select(log_salary, all_of(pred_vars)) %>%
  pivot_longer(-log_salary) %>%
  ggplot(aes(x = value, y = log_salary)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "Predictors vs log(Salary)")
```

### Correlation and Multicollinearity Check

```{r, warning=FALSE, message=FALSE}
# Correlation Plot / Matrix
cor_mat <- cor(df[, pred_vars], use = "complete.obs")
corrplot::corrplot(cor_mat, method = "color", addCoef.col = "black",
                   number.cex = 0.7, tl.cex = 0.8)

# Pairs plot (log Salary)
df %>%
  mutate(log_salary = log(Salary)) %>%
  select(log_salary, all_of(pred_vars)) %>%
  ggpairs()

# VIF
ols_fit <- lm(log(Salary) ~ USG + TO + TS + PpG + RpG + ApG + SpG + BpG,
              data = df)
vif(ols_fit)
```

We notice that for the heat map, there's no correlation coefficient greater than $0.8$, which means that all the variables at this stage are not considered to be highly correlated to each other. So we decide not to drop any of them.

Note that for our VIF values, there exists one VIF value of $6.016229$ (`PpG`) which is normally considered to be a concerning threshold that requires further investigation since this variable could potentially be affecting coefficient estimates. Since `PpG` represents points per game, higher `PpG` surely results in higher `Salary`. However, we decide not to drop this column, since this VIF value is not extremely high (less than 10), and this variable acts as a key factor to consider for our research question.

According to the distributions of our explanatory variables, we notice that those variables all have different ranges. In order to do later on analysis, we need to standardize our 8 predictors to make our prior specifications become consistent. Without standardization, coefficients are not comparable. In conclusion, Standardization ensures that the same prior is equally informative for all predictors regardless of their original units, and the posterior coefficients can be directly compared to assess the relative importance of each stat.

```{r}
df <- df %>%
  mutate(
    log_salary = log(Salary),
    USG_s = scale(USG)[,1],
    TO_s  = scale(TO)[,1],
    TS_s  = scale(TS)[,1],
    PpG_s = scale(PpG)[,1],
    RpG_s = scale(RpG)[,1],
    ApG_s = scale(ApG)[,1],
    SpG_s = scale(SpG)[,1],
    BpG_s = scale(BpG)[,1]
  )
head(df)
```

## Model 1: Bayesian Linear Regression

Our naive model is linear regression.

$$\log(salary)_i \sim \mathcal{N}(\alpha + \beta_1 \cdot USG_i + \beta_2\cdot TO_i + \beta_3\cdot TS_i + \beta_4 \cdot PpG_i + \beta_5 \cdot RpG_i + \beta_6 \cdot ApG_i + \beta_7 \cdot SpG_i +\beta_8 \cdot BpG_i,\ \sigma^2)$$

And we have

$$\alpha \sim \mathcal{N}(16, 2) \\ \beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8 \sim \mathcal{N}(0,1) \\ \sigma \sim \text{Half-Normal}(0, 1)$$

```{r}
## for checking
ols_fit <- lm(log(Salary) ~ USG_s + TO_s + TS_s + PpG_s + 
                RpG_s + ApG_s + SpG_s + BpG_s, data = df)
summary(ols_fit)$sigma
```

## Model 2: Hierarchical Bayesian Linear Regression

Our complex model is Hierarchical Linear Regression with position-specific intercepts.

$$\log(salary)_i \sim \mathcal{N}(\alpha_{j_[i]} + \beta_1 \cdot USG_i + \beta_2\cdot TO_i + \beta_3\cdot TS_i + \beta_4 \cdot PpG_i + \beta_5 \cdot RpG_i + \beta_6 \cdot ApG_i + \beta_7 \cdot SpG_i +\beta_8 \cdot BpG_i,\ \sigma^2)$$ Where $j[i]$ represents the i-th basketball player's position, $j = 1,2,...,7$, that we have different 7 positions in total.

$$\alpha_j \sim \mathcal{N}(\mu_\alpha, \tau^2), \ j = 1,2,...,7 \\ \mu_\alpha \sim \mathcal{N}(16,2) \\ \tau \sim \text{Half-Normal}(0, 1) \\ \beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8 \sim \mathcal{N}(0,1) \\ \sigma \sim \text{Half-Normal}(0, 1)$$

## Run Posterior Using Stan and R Code

```{r}
library(rstan)
library(loo)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

pred_vars <- c("USG", "TO", "TS", "PpG", "RpG", "ApG", "SpG", "BpG")

# 标准化预测变量（重要！让 beta ~ N(0,1) 真正 weakly informative）
X_raw <- df[, pred_vars]
X_scaled <- scale(X_raw)           # 保存 center / scale 属性，预测时还原用

y <- log(df$Salary)

# position index（整数编码）
pos_levels <- levels(factor(df$POS))
pos_id     <- as.integer(factor(df$POS, levels = pos_levels))
J          <- length(pos_levels)
N          <- nrow(df)
K          <- length(pred_vars)

cat("N =", N, "| K =", K, "| J =", J, "\n")
cat("Positions:", pos_levels, "\n")

# ── 1. Model 1：Bayesian Linear Regression ────────────────────
stan_data_m1 <- list(
  N = N,
  K = K,
  X = X_scaled,
  y = y
)

fit_m1 <- stan(
  file    = "model1.stan",
  data    = stan_data_m1,
  iter    = 4000,          # 2000 warmup + 2000 sampling
  warmup  = 2000,
  chains  = 4,
  seed    = 42,
  control = list(adapt_delta = 0.95)
)

print(fit_m1, pars = c("alpha", "beta", "sigma"), probs = c(0.025, 0.5, 0.975))

# ── 2. Model 2：Hierarchical Bayesian Linear Regression ───────
stan_data_m2 <- list(
  N      = N,
  K      = K,
  J      = J,
  X      = X_scaled,
  y      = y,
  pos_id = pos_id
)

fit_m2 <- stan(
  file    = "model2.stan",
  data    = stan_data_m2,
  iter    = 4000,
  warmup  = 2000,
  chains  = 4,
  seed    = 42,
  control = list(adapt_delta = 0.95)
)

print(fit_m2, pars = c("mu_a", "tau", "alpha", "beta", "sigma"),
      probs = c(0.025, 0.5, 0.975))

# ── 3. 收敛诊断 ───────────────────────────────────────────────
# R-hat 应全部 < 1.01，n_eff 应足够大
stan_diag_m1 <- summary(fit_m1)$summary
stan_diag_m2 <- summary(fit_m2)$summary

cat("\nModel 1 - max R-hat:", max(stan_diag_m1[, "Rhat"], na.rm = TRUE), "\n")
cat("Model 2 - max R-hat:", max(stan_diag_m2[, "Rhat"], na.rm = TRUE), "\n")

# ── 4. LOO-CV 比较两个模型 ────────────────────────────────────
log_lik_m1 <- extract_log_lik(fit_m1, merge_chains = FALSE)
log_lik_m2 <- extract_log_lik(fit_m2, merge_chains = FALSE)

loo_m1 <- loo(log_lik_m1, r_eff = relative_eff(exp(log_lik_m1)))
loo_m2 <- loo(log_lik_m2, r_eff = relative_eff(exp(log_lik_m2)))

# print(loo_compare(loo_m1, loo_m2))
cat("\nLoo compare:", loo_compare(loo_m1, loo_m2), "\n")

# ── 5. 后验预测：识别 overpaid / underpaid ────────────────────
# 用 Model 2（更优模型）的 posterior predictive distribution
y_rep_m2  <- extract(fit_m2, "y_rep")$y_rep   # shape: [draws, N]
residuals <- extract(fit_m2, "residual")$residual

# 每个球员：后验预测均值 & 95% 区间（log scale）
ppd_mean  <- apply(y_rep_m2, 2, mean)
ppd_lower <- apply(y_rep_m2, 2, quantile, 0.025)
ppd_upper <- apply(y_rep_m2, 2, quantile, 0.975)

results <- data.frame(
  NAME          = df$NAME,
  POS           = df$POS,
  Actual_Salary = df$Salary,
  log_Salary    = y,
  PPD_mean      = ppd_mean,
  PPD_lower     = ppd_lower,
  PPD_upper     = ppd_upper,
  # 换回原始薪资尺度
  Predicted_Salary = exp(ppd_mean),
  Lower_Salary     = exp(ppd_lower),
  Upper_Salary     = exp(ppd_upper)
)

# 判定：实际 log salary 超出 95% 后验预测区间
results$Status <- "Fair"
results$Status[results$log_Salary > results$PPD_upper] <- "Overpaid"
results$Status[results$log_Salary < results$PPD_lower] <- "Underpaid"

cat("\nOverpaid players:\n")
print(results[results$Status == "Overpaid",
              c("NAME", "POS", "Actual_Salary", "Predicted_Salary")])

cat("\nUnderpaid players:\n")
print(results[results$Status == "Underpaid",
              c("NAME", "POS", "Actual_Salary", "Predicted_Salary")])

```

```{r}
library(rstan)
library(ggplot2)
library(dplyr)

# ── 提取两个模型的 y_rep ──────────────────────────────────────
y_rep_m1 <- extract(fit_m1, "y_rep")$y_rep   # shape: [8000, 531]
y_rep_m2 <- extract(fit_m2, "y_rep")$y_rep   # shape: [8000, 531]

# ── 计算每个球员的 PPD 均值和 95% CI ─────────────────────────
ppd_summary <- data.frame(
  NAME          = df$NAME,
  POS           = df$POS,
  Actual_logSalary = log(df$Salary),
  Actual_Salary    = df$Salary,

  # Model 1
  M1_mean  = apply(y_rep_m1, 2, mean),
  M1_lower = apply(y_rep_m1, 2, quantile, 0.025),
  M1_upper = apply(y_rep_m1, 2, quantile, 0.975),

  # Model 2
  M2_mean  = apply(y_rep_m2, 2, mean),
  M2_lower = apply(y_rep_m2, 2, quantile, 0.025),
  M2_upper = apply(y_rep_m2, 2, quantile, 0.975)
)

# 换回原始薪资尺度
ppd_summary <- ppd_summary %>%
  mutate(
    M1_mean_salary  = exp(M1_mean),
    M1_lower_salary = exp(M1_lower),
    M1_upper_salary = exp(M1_upper),
    M2_mean_salary  = exp(M2_mean),
    M2_lower_salary = exp(M2_lower),
    M2_upper_salary = exp(M2_upper)
  )

# ── 打印数值对比表（log scale）────────────────────────────────
cat("PPD Credible Interval Comparison (log salary scale):\n")
cat(sprintf("%-20s %-5s %8s | %8s %8s %8s | %8s %8s %8s\n",
            "NAME", "POS", "Actual",
            "M1_mean", "M1_lower", "M1_upper",
            "M2_mean", "M2_lower", "M2_upper"))
cat(strrep("-", 95), "\n")

for (i in 1:nrow(ppd_summary)) {
  cat(sprintf("%-20s %-5s %8.3f | %8.3f %8.3f %8.3f | %8.3f %8.3f %8.3f\n",
              ppd_summary$NAME[i],
              ppd_summary$POS[i],
              ppd_summary$Actual_logSalary[i],
              ppd_summary$M1_mean[i],  ppd_summary$M1_lower[i],  ppd_summary$M1_upper[i],
              ppd_summary$M2_mean[i],  ppd_summary$M2_lower[i],  ppd_summary$M2_upper[i]))
}

# ── 可视化：随机抽30个球员展示 PPD CI 对比 ───────────────────
set.seed(42)
sample_players <- sample(ppd_summary$NAME, 30)
plot_df <- ppd_summary %>%
  filter(NAME %in% sample_players) %>%
  arrange(Actual_logSalary)

# 整理成长格式方便 ggplot
plot_long <- bind_rows(
  plot_df %>% mutate(model = "Model 1", mean = M1_mean, lower = M1_lower, upper = M1_upper),
  plot_df %>% mutate(model = "Model 2", mean = M2_mean, lower = M2_lower, upper = M2_upper)
)

ggplot(plot_long, aes(x = reorder(NAME, Actual_logSalary), color = model)) +
  geom_point(aes(y = mean), position = position_dodge(width = 0.6), size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                position = position_dodge(width = 0.6), width = 0.4) +
  geom_point(data = plot_df, aes(x = NAME, y = Actual_logSalary),
             color = "black", shape = 4, size = 3, inherit.aes = FALSE) +
  coord_flip() +
  labs(
    title    = "PPD 95% Credible Intervals: Model 1 vs Model 2",
    subtitle = "Black cross = actual log salary",
    x        = "Player",
    y        = "log(Salary)",
    color    = "Model"
  ) +
  theme_bw()

# ── overpaid / underpaid 判定（基于各自模型的 95% CI）────────
ppd_summary <- ppd_summary %>%
  mutate(
    M1_status = case_when(
      Actual_logSalary > M1_upper ~ "Overpaid",
      Actual_logSalary < M1_lower ~ "Underpaid",
      TRUE ~ "Fair"
    ),
    M2_status = case_when(
      Actual_logSalary > M2_upper ~ "Overpaid",
      Actual_logSalary < M2_lower ~ "Underpaid",
      TRUE ~ "Fair"
    )
  )

cat("\n── Model 1 status summary ──\n")
print(table(ppd_summary$M1_status))

cat("\n── Model 2 status summary ──\n")
print(table(ppd_summary$M2_status))

cat("\n── Players where two models disagree ──\n")
disagree <- ppd_summary %>%
  filter(M1_status != M2_status) %>%
  select(NAME, POS, Actual_Salary, M1_status, M2_status,
         M1_mean_salary, M2_mean_salary)
print(disagree)
```
